Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations in the Upper Salmon River, Idaho

### 2021-12-20 analysis/paper/paper.Rmd
edits to abundance & gastric methods

One potentially important, but perhaps under-appreciated source of salmonid mortality is predation on emigrating juveniles by piscivorous fishes. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of population recover [@Fresh2003; @Winther2019]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow water habitats [@Wydoski2003] and consequently have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Indeed, Northern Pikeminnow are estimated to consume 8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas very little is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately 1.5 kilometer long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is located in a river reach containing historically important overwinter rearing and spring emigration habitat for juvenile Chinook Salmon *O. tshawytscha* and is inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*. Further, the slough is part of the migratory pathway of the endangered Snake River Sockeye Salmon population [@Axel2015] and several populations of threatened Snake River steelhead.

@@

Our first objective was to estimate the population size of piscine predators in the Deadwater Slough using mark-recapture methods. We attempted to sample predators near the peaks of the juvenile DSR and NRR emigrations for the 2018 brood year (fall 2019 and spring 2020, respectively). However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 effort was moved to fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional effort was added in spring 2021. Each of the fall efforts included sampling across a two-week span; however, due to budget constraints the spring 2021 effort only included a single week. For the fall efforts, the first week was considered a "mark event" and the second week a "recapture event" for single-census mark-recapture estimators (described below); therefore, the "recapture week" occurred the week following the "mark week" to minimize predator emigration/immigration and constrain the population for the mark-recapture models.

@@

Multiple capture methods were employed during the fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. However, some methods proved ineffective during initial efforts. For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. After evaluating all methods, angling proved to be the most effective method for capturing piscine predators while minimizing potential impacts to steelhead but recognizing that angling can have both species and size selectivity (bias). From here on, we focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were permitted to fish anywhere within Deadwater Slough. During a sample event, anglers would boat or hike their catch (periodically or upon filling a livewell) to a processing station located at the boat ramp approximately 500 m downstream from the top of the slough. For each fish, we recorded the date of capture, species, total length (TL; mm), and whether the fish was previously marked. Unmarked fish were given a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique to each day and released. Released individuals were distributed throughout Deadwater Slough to facilitate mixing back into the population. Finally, we recorded the angling start and end times for each crew (person or combinations of persons) to facilitate calculations of catch per unit effort (CPUE).

@@

@@

Estimators fell into two broad categories: single census and multiple census. For the single census estimators, we treated the first week of sampling as the mark event, and the following (second) week as the recapture event, pooling data within each of those two weeks. Alternatively, the multiple census estimators treat each day as a survey, and use information about the total marked fish from all previous surveys to infer the total abundance.

The Lincoln-Peterson estimator, a single-census estimator, is below, where $M$ is the number of fish marked and returned to the population, the $n$ is the number of fish caught in the second/recapture event, and $m$ is the number of marked fish in the second sample.

The Lincoln-Petersen estimator can be biased with small samples, so we also employed the Chapman-modified Lincoln-Petersen estimator:

For a multiple-census estimator, we first utilized the Schnabel estimator where the $M$, $n$ and $m$ are indexed by the sampling occasion, $i$. In our case, the sampling occassions are defined as each day of sampling. The Schnabel estimator does not have an associated standard error; however, the 90% confidence intervals can be calculated. The Schnabel estimator is essentially a weighted average of a series of Lincoln-Peterson estimators (with a Chapman modfication). The Schnabel estimator is shown here:

Finally, we explored one additional estimator for multiple census surveys, called the Schumacher-Eschmeyer estimator, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and fall 2020 sampling periods. However, due to logistical constraints, we were only able to sample during a single week in spring 2021 and released fish were left unmarked. Instead, to estimate the abundance of Northern Pikeminnow during spring 2021, we used the ratio of total CPUE in the fall sampling events to the total CPUE in the spring sampling event, and multiplied that ratio by the average abundance of Northern Pikeminnow from the fall sampling events.<!-- Kevin, is this previous statement accurate? --> This approach assumes equal capture probabilities between the fall and spring capture events. If that assumption is not true and capture probability did differ, we believe that the capture probability  in the spring would be lower due to higher flows, which would lead to our spring abundance estimate to be biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 sampling event to be a conservative estimate.

```{r psd}
s_size = 300
q_size = 400

@@

Finally, we calculated the proportional stock density (PSD) for Northern Pikeminnow captured during all field efforts, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality-sized" individuals or larger individuals and "stock-sized" individuals. Stock and quality size definitions vary by species. PSD is expressed as:

@@

where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$. We calculated the PSD for Northern Pikeminnow in our study using `r prettyNum(s_size, digits = 0)` mm TL for stock-size and `r prettyNum(q_size, digits = 0)` mm TL for quality-size.

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner *Richardsonius balteatus*, etc.). Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks for later analysis in a controlled environment. For each sample, wet weight (grams) was recorded for the total combined stomach content, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as the portion consisting of fish parts/matter. Fish and fish remnants were identified to the lowest taxonomic unit, when possible, or were categorized as unknown. A subset of Northern Pikeminnow captures (~ 5%) were euthanized for dissection after gastric lavage to validate the efficacy of the methodology.

>>>>

One potentially important, but perhaps under-appreciated source of salmonid mortality is predation on emigrating juveniles by piscivorous fishes. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of population recover [@Fresh2003; @Winther2020]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow water habitats [@Wydoski2003] and consequently have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Indeed, Northern Pikeminnow are estimated to consume 8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas very little is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately 1.5 kilometer long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is located in a river reach containing historically important overwinter rearing and spring emigration habitat for juvenile Chinook Salmon *O. tshawytscha* and is inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*. Further, the slough is part of the migratory pathway of the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead.

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the juvenile DSR and NRR emigrations. Our intent was to sampling during the emigration of the 2018 brood year (fall 2019 and spring 2020). However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 effort was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional effort was added in spring 2021. Fall efforts included sampling across a two-week span; however, due to budget constraints the spring 2021 effort was conducted over a single week. Sampling was constrained within a two-week window to minimize predator emigration/immigration and to constrain the population for the mark-recapture models.

@@

Multiple capture methods were employed during the fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. However, ineffective methods were abandoned in subsequent years. After evaluating all methods, angling proved to be the most effective method for capturing piscine predators while minimizing potential impacts to ESA-listed steelhead, but recognizing that angling can have both species and size selectivity (bias). From here on, we focus on fish captured by angling, unless otherwise noted. Additionally, it quickly became clear during fall 2019 that Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and so is the focus of our study.

<!-- For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. -->

Our study relied heavily on volunteer anglers who were permitted to fish anywhere within Deadwater Slough. During a sampling occassion (day), anglers would boat or hike their catch (periodically or upon filling a livewell) to a processing station at the boat ramp approximately 500 m downstream from the top of the slough. For each fish, we recorded the date of capture, species, total length (TL; mm), and whether the fish was previously marked. Unmarked fish were given a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique to each day and released. Fish that died prior to release were still counted as captured (or recaptured), but not included in the number of marked fish available for recapture after that date. Released individuals were distributed throughout Deadwater Slough to facilitate mixing back into the population. Finally, we recorded the angling start and end time for each crew (person or combinations of persons) to facilitate calculations of CPUE.

@@

Mark-recapture abundance estimators fell into two broad categories: single-census and multiple-census. For the single-census estimators, we used the Lincoln-Peterson estimator:

where $M$ is the number is the total number of fish marked and returned to the population during the first week of sampling (mark event), $n$ is the total number of fish caught in the second week (recapture event), and $m$ is the number of marked fish caught during the recapture event.

Because the Lincoln-Peterson estimator can be biased with small samples, we also employed the Chapman-modified Lincoln-Peterson estimator:

For a multiple-census estimator, we first utilized the Schnabel estimator:

where $M$, $n$, and $m$ are indexed by the sampling occassion, $i$. Here, $i$ is defined as each sampling day. The Schnabel estimator does not have an associated standard error; however, the 90% confidence intervals were calculated.

Finally, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer estimator, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and fall 2020 sampling periods. However, due to logistical constraints, we were only able to sample during a single week in spring 20201 and released fish were left unmarked. Instead, to estimate the abundance of Northern Pikeminnow during spring 2021, we used the ratio of total CPUE in the fall sampling events to the total CPUE in the spring sampling event, and multiplied that ratio by the average abundance of Northern Pikeminnow from the fall sampling events.<!-- Kevin, is this previous statement accurate? --> This approach assumes equal capture probabilities between the fall and spring capture events. If that assumption is not true and capture probability did differ, we believe that the capture probability  in the spring would be lower due to higher flows, which would lead to our spring abundance estimate to be biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 sampling event to be a conservative estimate.

@@

```{r psd}
s_size = 250
q_size = 380

@@

<!--
Finally, we calculated the proportional stock density (PSD) for Northern Pikeminnow captured during all field efforts, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality-sized" individuals or larger individuals and "stock-sized" individuals. Stock and quality size definitions vary by species. PSD is expressed as:

@@

where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$. We calculated the PSD for Northern Pikeminnow in our study using `r prettyNum(s_size, digits = 0)` mm TL for stock-size and `r prettyNum(q_size, digits = 0)` mm TL for quality-size [@Winther2020].
-->

<!--
Juvenile Chinook Salmon abundances were estimated at 5 rotary screw traps in the upper Salmon River basin: Sawtooth, Pahsimeroi, Hayden Creek, Upper Lemhi, and Lower Lemhi. Abundances were calculated by Idaho Department of Fish and Game at each trap which made up the production estimate for juvenile Chinook Salmon in the Upper Salmon Basin (Table X) (McClure et al. 2020).
-->

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner *Richardsonius balteatus*, etc.). Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks for later analysis in a controlled environment. For each sample, wet weight (grams) was recorded for the total combined stomach content, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as the portion consisting of fish parts/matter. Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones, or were categorized as unknown [@Hansel1988; @Frost2000]. A subset of Northern Pikeminnow captures (~ 5%) were euthanized for dissection after gastric lavage to validate the efficacy of the methodology.
